---
title: "Califano Master Regulator Discovery Analyses"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


### REQUIREMENTS
+ JAVA RUNTIME ENVIRONMENT

### BACKGROUND
+ [The recurrent architecture of tumour initiation, progression and drug sensitivity](https://www.nature.com/articles/nrc.2016.124)
+ [ARACNE-AP](https://github.com/califano-lab/ARACNe-AP)


### SPECIFY INPUT DATA AND PARAMETERS
```{r setup}
# NOTE: FOR ACCURATE ESTIMATION OF THE PAIRWISE (GENE-GENE)
#       MUTUAL INFORMATION MEASURE USED FOR CONTEXT-SPECIFIC 
#       NETWORK RECONSTRUCTION, THE EXPRESSION MATRIX SHOULD HAVE
#       **100 OR MORE** SAMPLES (SPANNING A BROAD RANGE OF CELLULAR STATES
#       AND EXPRESSION LEVELS WITHIN A RELATED TUMOR/TISSUE CONTEXT).
#       
# ******************************************************************************
# GENE EXPRESSION DATA FILE
# Expected format: tab_separated data table (genes x samples)
# --- column 1 (column name = gene) lists genes
# --- other columns: named by sample identifier
exp_data_file <- "data-raw/tcga_gbm_exp.txt"

# TRANSCRIPTION FACTOR LIST FILE
# Expected format: Text file - one gene symbol per line, no column name/header.
# --- Check: entries in this file are named in column 1 of above exp data table.
# --- Check: https://www.genecards.org/ for alternative gene names (as needed).
tflist_file <- "data-raw/tcga_gbm_mesenchymal_transfacs.txt"

output_dir <- "TCGA_GBM_ARACNE"

# p-value for mutual information to be significant
# lower values reduce number of associations in in network.
aracne_mutinf_pval_threshold <- 1E-8

# Final network is consensus over networks constructed from bootstrapped
# sample expression subsets.
n_aracne_bootstraps <- 100
# ******************************************************************************

library(tidyverse)

```

### (0) Prepare Expression Data
```{r prep_expression_data}
transfacs <- readr::read_tsv(tflist_file, show_col_types = FALSE, 
                             col_names = FALSE)$X1
exp_tbl <- readr::read_tsv(exp_data_file, show_col_types = FALSE)

# PREPROCESSING: remove genes w/expression value of zero in all samples --------
exp_data_file <- paste0(
  tools::file_path_sans_ext(exp_data_file),
  ".reduced.",
  tools::file_ext(exp_data_file)
)

exp_mat <- exp_tbl %>%
  as.data.frame() %>%
  column_to_rownames(var = "gene") %>%
  as.matrix()
stopifnot(is.numeric(exp_mat))

all_zero_exp <- apply(exp_mat, MARGIN = 1, FUN = function(x) {
  all(dplyr::near(x, 0))
})
exp_mat <- exp_mat[!all_zero_exp, , drop = FALSE]

exp_tbl <- exp_mat %>%
  as.data.frame() %>%
  rownames_to_column(var = "gene") %>%
  as_tibble()

rm(exp_mat)

readr::write_tsv(exp_tbl, file = exp_data_file)

if (!all(transfacs %in% exp_tbl$gene)) {
  warning(
    "No available expression data for these TFs (check alternative names): ",
    paste0(setdiff(transfacs, exp_tbl$gene), collapse = " ")
  )
}

```


### (1) Estimate Mutual Information Threshold for ARACNE Network Reconstruction
```{r calculate_aracne_threshold}
# Sample command line:
# java -Xmx5G -jar ./ARACNe-AP/dist/aracne.jar \
#   -e ./data-raw/tcga_gbm_exp.reduced.txt  \
#   -o TCGA_GBM_ARACNE \
#   --tfs ./data-raw/tcga_gbm_mesenchymal_transfacs.txt \
#   --pvalue 1E-8 \
#   --seed 1 \
#   --calculateThreshold

system2(
  command = "java",
  args    = c(
    "-Xmx5G", "-jar", "./ARACNe-AP/dist/aracne.jar",
    "-e", exp_data_file,
    "-o", output_dir,
    "--tfs", tflist_file,
    "--pvalue", as.character(aracne_mutinf_pval_threshold),
    "--seed", "1",
    "--calculateThreshold"
  )
)
```


### (2) Bootstrapping/Network Reconstruction
```{r bootstrap_network_reconstruction}
# Sample command line (per iterate):
# java -Xmx5G -jar ./ARACNe-AP/dist/aracne.jar   \
#   -e ./data-raw/tcga_gbm_exp.reduced.txt    \
#   -o TCGA_GBM_ARACNE   \
#   --tfs ./data-raw/tcga_gbm_mesenchymal_transfacs.txt \
#   --pvalue 1E-8 \
#   --seed 1

for (i in seq_len(n_aracne_bootstraps)) {
  print(paste0("--------------- bootstrap iterate: ", i, "  ----------------"))
  
  system2(
    command = "java",
    args    = c(
      "-Xmx5G", "-jar", "./ARACNe-AP/dist/aracne.jar",
      "-e", exp_data_file,
      "-o", output_dir,
      "--tfs", tflist_file,
      "--pvalue", as.character(aracne_mutinf_pval_threshold),
      "--seed", as.character(i)
    )
  )
}

```


### (3) Build Consensus Regulatory Network
```{r}
#java -Xmx5G -jar ./ARACNe-AP/dist/aracne.jar \
#   -o TCGA_GBM_ARACNE \
#   --consolidate

system2(
    command = "java",
    args    = c(
      "-Xmx5G", "-jar", "./ARACNe-AP/dist/aracne.jar",
      "-o", output_dir,
      "--consolidate"
    )
  )
```


### Review Consensus Network Attributes
```{r review_network_attributes}
aracne_network_file <- file.path(output_dir, "network.txt")

aracne_network_tbl <- readr::read_tsv(aracne_network_file, 
                                      show_col_types = FALSE)

# Here we're supposing that the TFs in the transcription factor list file
# constitute a phenotypic regulatory checkpoint module and we're looking their
# interconnectivity.
checkpt_tbl <- aracne_network_tbl %>% filter((Target %in% transfacs)) %>%
  arrange(Regulator, pvalue)

quantile(aracne_network_tbl$MI, probs = c(0, 0.5, 0.75, 0.85, 0.90, 0.95, 1))

```

